---
title: "Statistics for data from Fig 2h Caillaud _et al_ PLoS Biology"
author: "Dan MacLean"
date: "13 August 2015"
output: word_document
---

## Pre-processing

Marie-Cecille Caillaud (MCC) sent me an Excel file of all the haemocytometry measurements she made - file `raw/MCC-Dan corrected.xslx`. The file annotates figures from the same biological replicates as colours which I can't parse programmatically. I therefore added columns to the sheet stating the replicate number. I also removed spaces in column headers and saved the file as `raw/MCC-Dan corrected Reps added.xlsx` and exported the sheet with the data to a csv file `fig_2h_data_manual.csv` which I can operate on programmatically and will use as my input.

### Use some Python to get the data file into better shape
```{python}
header = []
results = []
with open('raw/fig_2h_data_manual.csv', 'r') as file:
  for l in file:
    l = l.rstrip('\r\n')
    a = l.split(',')
    if l.startswith("Rep"):
      header = a
    else:
      for i in range(0,len(header),2):
        rep,line,count = a[i],header[i+1],a[i+1]
        if rep and line and count: ## if we have no empty values
          results.append([rep,line,count])

with open('data/reshaped_data.csv','w') as outfile:
  outfile.write("Replicate,Line,Count\n")
  for r in results:
    outfile.write(",".join(r) + "\n") 
```

### Load data, reorder as per Figure 2H and do a straightforward plot
```{r}
library(ggplot2)
data <- read.csv('data/reshaped_data.csv', header=TRUE)
data$Replicate <- as.factor(data$Replicate)
data$Line <- factor(data$Line, c("Col-0","Med15", "Med19-1","C1","C2","Med19a-2","GFP", "OE_med19-1","OE_med19-2","Med25","med21"))
basic <- ggplot(data, aes(Line,Count))
scatter <- basic + geom_jitter(aes(colour=Replicate),position = position_dodge(width=0.5)) + theme(axis.text.x = element_text(angle = 90, hjust = 1))
scatter
```
```{r}
data
```
The data look ok, a few outliers in `Med19a-2` and `C1` that could affect summary statistics. Let's do some `qqplots` and see how  
they lie.
```{r}
#qnorm is default distribution - we are testing for a normal distribution
ggplot(data, aes(sample=Count)) + geom_jitter(stat="qq", aes(colour=Replicate) ) + facet_wrap( ~ Line, scales = "free_y")
```

Those outliers could mess up summary statistics, they're off the curve, we have no good reason to ditch them though. I suppose they mean that occasionally the method used (spore counting) throws up some very extreme numbers. Overall these plots are ok, the variation seems normally distributed on the whole.

Let's have a look at summary statistics:
```{r}
library(plyr)
summary <- ddply(data,"Line",summarise, mean=mean(Count),median=median(Count),diff=abs(mean(Count) - median(Count)), std_dev=sd(Count), std_err=sd(Count)/sqrt(length(Count)))
summary
```

The summary stats seem fine overall, similar SD and SE and not much drift of the median from the mean, the concern again is `Med19a-2` and `C1` with the high standard deviation and mean dragged up by that couple of points.

## Does a bar chart imply a higher effect than we see generally?
Let's make a bar graph with error bars on that first scatter to see how using a standard bar chart might be misleading our thinking. 

```{r}
ggplot(summary, aes(x=Line, y=mean)) + geom_bar(position=position_dodge(), stat="identity", fill="steelblue",colour="black") + geom_errorbar(aes(ymin=mean-std_err, ymax=mean+std_err)) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) 
```

The barchart is definitely suggesting a higher overall effect than we see from the individual replicates in the scatter plot for `Med19a-2` and `C1` My conclusion here is that although the mean is calculated correctly, it's just that the mean is a slightly misleading number to boil our data down to in this case. Also that very slight increase in standard error isn't giving us a clue as to that messy single outlier. Taken together the mean and SE plotted like this convince of us a bigger effect in general so the plot style isn't helpful. 

## Significance Tests
I need to boil down the data to the biological replicates.
```{r}
library(reshape)
bioreps <- cast(data, Line~Replicate, mean)
bioreps <- melt(bioreps)
bioreps
```

I'll do an ANOVA and Tukey's HSD for multiple comparisons.
```{r}
### ANOVA and Tukey's HSD on all pairwise - though really only interested in VS Col-0 the control
fit <- aov(lm(value ~ Line,data=bioreps))
TukeyHSD(fit)               
```

A long table, but it's showing the overexpressers `OE_med19-1` and `OE_med19-2` are different from the `Col-0` control, as is the one with the noted high outliers `Med19a-2` and also `Med15`.

## P-Hacking
Let's see how removing those high (`>=150`) outliers affects the _p_-values, see if any signficance we have is coming from one or two atypical data.

```{r}
under_150 <- data[data$value < 150, ]
bioreps_under150 <- cast(data, Line~Replicate, mean)
bioreps_under150 <- melt(bioreps_under150)
fit <- aov(lm(value ~ Line,data=bioreps_under150))
TukeyHSD(fit)
```

Looks good! The same Lines come up as significant - the outliers aren't messing with the overall significance result.

### More P-Hacking - ditching data originally in Figure 2H!

According to MCC and JJ then the lines of interest are really the `med19-1`, `Med19a-2`, `OE_med19-1` and `OE_med19-2`. Let's do the same tests for the restricted set and see if it substantially affects the result.
```{r}
of_interest <- bioreps_under150[bioreps_under150$Line %in% c("Col-0", "Med19-1", "Med19a-2", "OE_med19-1", "OE_med19-2"), ]
fit <- aov(lm(value ~ Line,data=of_interest))
TukeyHSD(fit)
```

The result is not substantially different from before, the same lines show up as significantly different, that is `Med19a-2`, `OE_med19-1`, `OE_med19-2` and `Med15` are signifcantly different from the `Col-0` control. `Med19-1` is not.

## Conclusion
The Med19-2 and Med15 lines get significantly more spores than the Col-0 wild-type and the two over-expressors of Med19 show significantly fewer spores than Col-0. There is no evidence for difference from the wild-type and other lines. 

```{r,echo=FALSE}
### Now produce a pretty version of the scatter plot and save to file
library(ggthemes)
pretty <- scatter + theme_few() + scale_colour_few() +  theme(axis.text.x = element_text(angle = 90, hjust = 1))
ggsave('figure.png', pretty)
```
